---
title: "Census Data in R, Part 2: Mapping with tidycensus and sf"
subtitle: "Pulling census geometry and making choropleth maps of North Carolina"
description: "Add geometry to your tidycensus requests and map census variables for North Carolina counties and Mecklenburg County census tracts."
order: 3
---

::: {.callout-note}
## About this tutorial

**Time:** about 25 minutes. **Prerequisites:** [Census Data in R, Part 1](census-r-part1.qmd), with your API key installed.

By the end, you will be able to:

- request census data with shapes attached
- make a choropleth map with ggplot2
- choose a map projection that suits North Carolina
- map census tracts, remove large bodies of water, and add county outlines
- make an interactive map for exploring your data
- attach census shapes to data from somewhere else
:::

## Setup

You need three packages beyond the ones from Part 1. [sf](https://r-spatial.github.io/sf/){target="_blank"} handles spatial data in R. [tigris](https://github.com/walkerke/tigris){target="_blank"} downloads Census Bureau boundary files. tidycensus uses tigris behind the scenes. [leaflet](https://rstudio.github.io/leaflet/){target="_blank"} makes interactive maps.

```{r}
#| eval: false
install.packages(c("sf", "tigris", "leaflet"))
```

```{r}
#| message: false
#| cache: false
library(tidycensus)
library(tidyverse)
library(sf)
library(tigris)

options(tigris_use_cache = TRUE)
```

The `tigris_use_cache` option saves boundary files on your computer after the first download. That makes later requests much faster.

## Adding geometry to a request

To get shapes along with your data, add `geometry = TRUE` to `get_acs()` or `get_decennial()`. Everything else stays the same:

```{r}
nc_income <- get_acs(
  geography = "county",
  state = "NC",
  variables = c(median_income = "B19013_001"),
  year = 2024,
  geometry = TRUE
)

nc_income
```

The result is now an **sf object**. It looks like a tibble with one extra column named `geometry`, which stores each county's outline. The header also reports a **CRS** (coordinate reference system). Census shapes arrive in NAD83 (EPSG code 4269), which stores locations as longitude and latitude.

You can make a quick map with base R's `plot()`:

```{r}
plot(nc_income["estimate"])
```

## A choropleth map with ggplot2

A **choropleth** map shades each area by the value of a variable. With ggplot2, `geom_sf()` draws the shapes and the `fill` aesthetic sets the colors:

```{r}
ggplot(nc_income, aes(fill = estimate)) +
  geom_sf(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(labels = scales::label_dollar()) +
  labs(
    title = "Median household income by county",
    subtitle = "North Carolina, 2020 to 2024 ACS 5-year estimates",
    fill = NULL,
    caption = "Source: US Census Bureau"
  ) +
  theme_void()
```

A few choices in that code are worth pointing out:

- `scale_fill_viridis_c()` uses a color scale that is easy to read for people with color vision deficiencies. It also prints well in grayscale.
- `scales::label_dollar()` formats the legend as currency.
- `theme_void()` removes the axes and gridlines. Latitude and longitude ticks rarely help a reader.

### Choosing a projection

Longitude and latitude describe points on a globe. Drawing them directly on a flat screen stretches the map. North Carolina is wide and short, so the distortion is mild, but it is still there. The state has its own projected CRS, called NAD83 / North Carolina (EPSG code 32119). You can switch to it in one line:

```{r}
nc_income_proj <- st_transform(nc_income, crs = 32119)
```

Plot `nc_income_proj` with the same ggplot code and compare. The shapes will look a little more like the state you know from official maps. Projected coordinates are measured in meters. That also makes them the right choice for measuring areas and distances.

## Mapping census tracts

Tract maps show variation inside a county. Here is median household income for Mecklenburg County tracts:

```{r}
meck_income <- get_acs(
  geography = "tract",
  state = "NC",
  county = "Mecklenburg",
  variables = c(median_income = "B19013_001"),
  year = 2024,
  geometry = TRUE
) |>
  st_transform(32119)

ggplot(meck_income, aes(fill = estimate)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(labels = scales::label_dollar(), na.value = "grey80") +
  labs(
    title = "Median household income by census tract",
    subtitle = "Mecklenburg County, 2020 to 2024 ACS 5-year estimates",
    fill = NULL
  ) +
  theme_void()
```

Setting `na.value` gives tracts with suppressed estimates their own gray color. Without it, those tracts could be mistaken for part of the scale.

### Removing water

Census tract boundaries include water. On the northern edge of Mecklenburg County, the tracts along Lake Norman extend into the lake. The map would look more familiar with the lake cut out. `erase_water()` from tigris removes large bodies of water from your shapes:

```{r}
meck_income_dry <- erase_water(meck_income, area_threshold = 0.9)
```

The `area_threshold` argument controls how much water gets erased. A value of `0.9` removes only the largest 10 percent of water features. Run the map code again with `meck_income_dry` to see the difference. This step can take a minute the first time you run it.

### Mapping a region with county outlines

To show tracts across several counties, pass a vector of counties. Then draw county boundaries on top so readers can orient themselves:

```{r}
region_counties <- c("Mecklenburg", "Iredell", "Cabarrus", "Lincoln", "Gaston")

region_income <- get_acs(
  geography = "tract",
  state = "NC",
  county = region_counties,
  variables = c(median_income = "B19013_001"),
  year = 2024,
  geometry = TRUE
) |>
  st_transform(32119)

region_outlines <- counties(state = "NC", cb = TRUE, year = 2024) |>
  filter(NAME %in% region_counties) |>
  st_transform(32119)

ggplot() +
  geom_sf(data = region_income, aes(fill = estimate), color = NA) +
  geom_sf(data = region_outlines, fill = NA, color = "white", linewidth = 0.6) +
  scale_fill_viridis_c(labels = scales::label_dollar(), na.value = "grey80") +
  labs(title = "Median household income around Lake Norman", fill = NULL) +
  theme_void()
```

The `cb = TRUE` argument requests *cartographic boundary* files. These are simplified shapes that draw quickly and follow the coastline. tidycensus uses them by default when you set `geometry = TRUE`.

## Mapping several variables at once

Long-format data pairs nicely with `facet_wrap()`. This example maps four racial and ethnic groups across Mecklenburg tracts from the 2020 Redistricting file:

```{r}
#| results: hide
meck_race <- get_decennial(
  geography = "tract",
  state = "NC",
  county = "Mecklenburg",
  variables = c(
    White = "P1_003N",
    Black = "P1_004N",
    Asian = "P1_006N",
    Hispanic = "P2_002N"
  ),
  summary_var = "P1_001N",
  year = 2020,
  sumfile = "pl",
  geometry = TRUE
) |>
  mutate(percent = 100 * value / summary_value) |>
  st_transform(32119)

ggplot(meck_race, aes(fill = percent)) +
  geom_sf(color = NA) +
  facet_wrap(~variable) +
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "grey80") +
  labs(
    title = "Race and ethnicity by census tract, Mecklenburg County",
    subtitle = "2020 decennial census",
    fill = "Percent"
  ) +
  theme_void()
```

The four panels share one color scale. That makes it easy to compare where each group lives. One caution applies here. The Census Bureau counts Hispanic origin separately from race. A Hispanic resident may also be counted in one of the race groups, so the four percentages can add up to more than 100.

## Interactive maps

A static map is best for a presentation. While you explore, an interactive map lets you zoom in and hover over tracts to read their values. [leaflet](https://rstudio.github.io/leaflet/){target="_blank"} builds interactive web maps from R. It uses the same JavaScript library that powers many of the maps you see on news sites.

leaflet expects longitude and latitude, so first transform the data to EPSG 4326 (WGS84), the standard CRS for web maps. Then build a color palette that maps each income to a color:

```{r}
library(leaflet)

meck_web <- st_transform(meck_income, 4326)

income_pal <- colorNumeric("viridis", domain = meck_web$estimate)
```

A leaflet map is built in layers, much like a ggplot. Start with a basemap, then add the tracts, then add a legend:

```{r}
leaflet(meck_web) |>
  addProviderTiles(providers$Esri.WorldGrayCanvas) |>
  addPolygons(
    fillColor = ~income_pal(estimate),
    fillOpacity = 0.7,
    color = "white",
    weight = 0.5,
    label = ~paste0(NAME, ": ", scales::dollar(estimate), " ± ", scales::dollar(moe))
  ) |>
  addLegend(
    pal = income_pal,
    values = ~estimate,
    title = "Median household income",
    labFormat = labelFormat(prefix = "$")
  )
```

Here is what each layer does:

- `addProviderTiles()` draws a basemap underneath your data. Esri's World Gray Canvas is a light gray map with only faint roads and borders, so it doesn't compete with your colors. It is free and needs no API key. Some other providers, such as CARTO, now require a key. For a full street map, use `addTiles()` instead, which draws OpenStreetMap.
- `addPolygons()` draws the tracts. The `~` in front of an argument tells leaflet to look up a column in your data. `label` sets the text that appears when you hover over a tract.
- `addLegend()` explains the colors. It uses the same palette function as the polygons, so the two always match.

Hover over the tracts to see each estimate and its margin of error. This is a quick way to find the tracts behind a pattern you notice.

## Attaching census shapes to other data

Sometimes your data doesn't come from the Census Bureau. It might be a table of eviction filings, school test scores, or health outcomes by tract. If the table has a GEOID column, you can download shapes with tigris and join them:

```{r}
#| eval: false
meck_tracts <- tracts(state = "NC", county = "Mecklenburg", cb = TRUE, year = 2024)

# Suppose my_data has a GEOID column stored as text
meck_joined <- meck_tracts |>
  left_join(my_data, by = "GEOID")
```

Put the sf object on the left side of the join. The result then stays an sf object, and ggplot2 can map it. Before you join, check that the GEOIDs in both tables are text and that they have the same number of digits.

::: {.callout-warning}
## Match your years
Tract boundaries changed in 2020. Data that uses 2010 tract GEOIDs won't join cleanly to 2020 or later shapes. If many rows fail to match, check which decade's tracts your data uses.
:::

## Saving maps and shapes

Save a ggplot map as an image for your slides:

```{r}
#| eval: false
ggsave("meck_income_map.png", width = 8, height = 6, dpi = 300)
```

Save the spatial data itself as a GeoPackage. A GeoPackage keeps the shapes, the data, and the CRS together in one file. Teammates can open it in R, in Python, or in QGIS.

```{r}
#| eval: false
st_write(meck_income, "meck_income.gpkg")
```

## Try it yourself

1. Make a county map of the percentage of people below the poverty level in North Carolina, using the `summary_var` approach from Part 1. Project it to EPSG 32119.
2. Map the median gross rent (`B25064_001`) for census tracts in Mecklenburg, Iredell, and Cabarrus counties. Remove the water and add county outlines. Where are rents highest?
3. Make a second map that shows the **margin of error** (the `moe` column) for the same rent data. Which areas have the least reliable estimates? How would you mention that in a presentation?

## Further reading

From Kyle Walker's [*Analyzing US Census Data*](https://walker-data.com/census-r/){target="_blank"}:

- [Chapter 5: Census geographic data and applications in R](https://walker-data.com/census-r/census-geographic-data-and-applications-in-r.html){target="_blank"}, which covers tigris, projections, and cartographic boundary files
- [Chapter 6: Mapping Census data with R](https://walker-data.com/census-r/mapping-census-data-with-r.html){target="_blank"}, which covers ggplot2, tmap, and interactive maps in much more depth

**Using Python instead?** See [Census Data in Python, Part 1](census-python-part1.qmd).
